Keep your Laziness in Check

### 2018-03-09 paper/main.tex

How, then, can we test and catch such laziness bugs during
development? Is it possible to precisely specify laziness behavior
just as we would for functional correctness?
%
In this paper, we demonstrate techniques for observing laziness in
Haskell, and propose a system of creating specifications of lazy
Haskell programs. We implement these ideas as a Haskell library. For
concreteness, consider the following example using the standard
\lk{take} function on lists, which accepts an integer $n$ and a list, and returns
a prefix of the list with length $n$. If $n$ is larger than the length
of the original list, then the entire list is returned.
%
\begin{inlinecode}
take_spec :: Shaped a => Spec '[Int, [a]] [a]
take_spec =
  Spec $ \predict d n xs ->
    predict n (if n > length xs then d else d ++ thunk)
\end{inlinecode}
%
The \lk{Spec} constructor takes a function, whose first
argument \lk{predict} is the predicted demand on the \lk{Int} argument
of \lk{take}, and second argument the predicted demand on the \lk{[a]}
argument. This spec simply says that the \lk{Int} argument will always
be used, and the demand on the list argument is either the same as the
demand on the output, or demand on the output plus some
unevaluated \lk{thunk} at the end. The former case happens when $n$ is
larger than the list's length, otherwise the latter case applies.

In the next section we work through two more examples, functional
queues and functional maps, demonstrating both the difficulties of
programming with laziness and how our approach can be used to
alleviate them.
%
In the rest of the paper, we present our technical contributions:
\begin{itemize}
\item We introduce a novel approach to specifying the laziness behavior
of functions that is {\em precise} and {\em compositional}. Such
specifications are expressive enough to capture exactly how much
evaluation occurs within a data structure
%\leo{Also add
%drawbacks/partiality here? If not here, where in the intro?} \hzh{I'm
%not sure if we should bring in partiality (in the sense of
%divergence/nontermination) yet since that seems like a rather
%technical detail of the spec}
\item We propose a framework for automated testing against such specifications.\caleb{Automated testing? Can be more specific}\leo{Should we mention QuickCheck here?}
\item We improve upon the random function generators from QuickCheck\cn to generate functions with arbitrary laziness behaviors.
\item We discuss our generic programming-based impementation to apply StrictCheck to all algebraic data types.\leo{I guess ``packed'' maps do not qualified as algebraic?}
\end{itemize}
%
Section~\ref{sec:related} discusses related work. We conclude and draw
directions for future work in Section~\ref{sec:concl}.

@@

We can now move on to the more complex \lk{rot} function from Okasaki:
as long as the front list is longer than the back list, demanding the
first element from the result of \lk{rotate} only causes evaluation of
the first element of the front list ~\cite{Okasaki95simpleand}.
%
\begin{inlinecode}
rot_spec :: Shaped a => Spec '[[a], [a]] [a]
rot_spec =
  Spec $ \predict d fs bs ->
    let (fs', bs') = splitAt (length fs) (cap d)
        spineLen  = length (cap (d ++ [undefined]))  -- # of spine thunks forced
        overflow  = spineLen       > length fs  -- begun taking from bs?
        overrot   = length (cap d) > length bs  -- forced all of bs?
        padLength =
          length bs `min`
            if overflow
            then length bs - length bs'
            else length (cap d)
        spinePad = replicate padLength thunk
    in predict
         (                    fs' ++ if overflow            then [] else thunk)
         (spinePad ++ reverse bs' ++ if overflow || overrot then [] else thunk)
\end{inlinecode}
%
The \lk{rotate\_spec} exactly captures the core idea from Okasaki's \lk{rotate}
operation, and testing the reverse and append implementation against this
specification immediately produces a counter example.
\begin{figure}[H]
\centering
\includegraphics[width=0.8\textwidth]{revappend_counterexample}
\caption{Counterexample for \lk{revAppend} under \lk{rot\_spec}}
\end{figure}

>>>>

How, then, can we test and catch such laziness bugs during
development? Is it possible to precisely specify laziness behavior
just as we would for functional correctness?
%
In this paper, we demonstrate techniques for observing laziness in
Haskell, and propose a system of creating specifications of lazy
Haskell programs. We implement these ideas as a Haskell library. For
concreteness, consider the following example using the standard
\lk{take} function on lists, which accepts an integer $n$ and a list, and returns
a prefix of the list with length $n$. If $n$ is larger than the length
of the original list, then the entire list is returned.
%
\begin{inlinecode}
take_spec :: Shaped a => Spec '[Int, [a]] [a]
take_spec =
  Spec $ \predict d n xs ->
    predict n (if n > length xs then d else d ++ thunk)
\end{inlinecode}
%
The \lk{Spec} constructor takes a function, whose first
argument \lk{predict} is is a continuation that takes the predicted
demands on the inputs to \lk{take}, and second argument of the
function is the observed demand on the output \lk{[a]}. This spec says
that the \lk{Int} argument will always be used, and the demand on the
list argument is either the same as the demand on the output, or
demand on the output plus some unevaluated \lk{thunk} at the end. The
former case happens when $n$ is larger than the list's length,
otherwise the latter case applies.

In the next section we work through two more examples, functional
queues and functional maps, demonstrating both the difficulties of
programming with laziness and how our approach can be used to
alleviate them.
%
In the rest of the paper, we present our technical contributions:
\begin{itemize}
\item We introduce a novel approach to specifying the laziness behavior
of functions that is {\em precise} and {\em compositional}. Such
specifications are expressive enough to capture exactly how much
evaluation occurs within a data structure
%\leo{Also add
%drawbacks/partiality here? If not here, where in the intro?} \hzh{I'm
%not sure if we should bring in partiality (in the sense of
%divergence/nontermination) yet since that seems like a rather
%technical detail of the spec}
\item We propose a framework for automated testing against such specifications.\caleb{Automated testing? Can be more specific}\leo{Should we mention QuickCheck here?}
\item We improve upon the random function generators from QuickCheck ~\cite{Claessen:2000:QLT:357766.351266} to generate functions with arbitrary laziness behaviors.
\item We discuss our generic programming-based impementation to apply StrictCheck to all algebraic data types.\leo{I guess ``packed'' maps do not qualified as algebraic?}
\end{itemize}
%
Section~\ref{sec:related} discusses related work. We conclude and draw
directions for future work in Section~\ref{sec:concl}.

@@

We can now move on to the more complex \lk{rot} function from Okasaki:
as long as the front list is longer than the back list, demanding the
first element from the result of \lk{rotate} only causes evaluation of
the first element of the front list ~\cite{Okasaki95simpleand}.
%
\begin{inlinecode}
rot_spec :: Shaped a => Spec '[[a], [a]] [a]
rot_spec =
  Spec $ \predict d fs bs ->
    let (fs', bs') = splitAt (length fs) (cap d)
        spineLen  = length (cap (d ++ [undefined]))  -- # of spine thunks forced
        overflow  = spineLen       > length fs  -- begun taking from bs?
        overrot   = length (cap d) > length bs  -- forced all of bs?
        padLength =
          length bs `min`
            if overflow
            then length bs - length bs'
            else length (cap d)
        spinePad = replicate padLength thunk
    in predict
         (                    fs' ++ if overflow            then [] else thunk)
         (spinePad ++ reverse bs' ++ if overflow || overrot then [] else thunk)
\end{inlinecode}
%
The \lk{rot\_spec} exactly captures the core idea from Okasaki's \lk{rotate}
operation, and testing the reverse and append implementation against this
specification immediately produces a counter example.
\begin{figure}[H]
\centering
\includegraphics[width=0.8\textwidth]{revappend_counterexample}
\caption{Counterexample for \lk{revAppend} under \lk{rot\_spec}}
\end{figure}
